\documentclass[10pt]{article}

\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage[margin=1in]{geometry}
\usepackage[hidelinks]{hyperref}
\usepackage[expansion=false]{microtype}

\title{WAMProbe: Capability-Aware Counterfactual Evaluation\\for World Action Models}
\author{WAMProbe Contributors}
\date{Technical Report Draft, July 2026}

\begin{document}
\maketitle

\begin{abstract}
World Action Models (WAMs) combine future prediction with robot action modeling, but the
role of the predicted future is difficult to identify from task success or video fidelity
alone. We present WAMProbe, an open-source evaluation toolkit organized around paired
action interventions from a shared initial state. The toolkit routes metrics through an
explicit capability declaration and reports causal response, state accuracy, candidate
ranking, control regret, and compute cost as a profile rather than a composite score.
The current release candidate contains three analytic benchmarks, a four-family paired LIBERO
simulator pilot, and a pinned StarWAM action-prediction adapter. Reference baselines show
the intended failure signatures: a wrong-direction model responds to action while
predicting reversed dynamics, and action-ignoring models fail candidate ranking. A
controlled appearance corruption preserves exact state prediction and zero regret while
reducing PSNR to approximately 0.59 dB, which shows that pixel fidelity cannot replace
control-grounded measures. In an analytic score-execute-observe study, offline Candidate
Ranking Correlation has descriptive Pearson association 0.9855 and 1.0000 with mean
closed-loop return on BlockPush and Gripper-Catch, respectively, across five deliberately
separated future scorers. A separate 36-sample StarWAM matrix executes every predicted
action chunk successfully in LIBERO but obtains zero sparse task success at the tested
short horizons. We retain this negative result and do not infer real-model closed-loop
utility from the analytic study.
\end{abstract}

\section{Introduction}

Task success is necessary for robot policy evaluation, but it does not identify why a
system succeeds or fails. Recent WAM diagnostics report differences in motion
consistency, target progress, distractor motion, and runtime that disappear in a single
success number \cite{mai2026beyond}. A second ambiguity arises when a predicted video is
plausible but incompatible with the action that supposedly caused it
\cite{ruan2026compatible}. Explicit future generation can also add inference cost without
providing a corresponding control gain \cite{yuan2026fastwam}. These findings motivate a
more direct question: under the same observation, do different proposed actions lead to
the correct different futures, and do those futures select actions that improve control?

WAMProbe treats a shared initial state with multiple action branches as the primary
evaluation unit. This design supports paired comparisons and prevents context variation
from being mistaken for action sensitivity. It also exposes simple failure modes that a
single rollout hides. A copy-last predictor ignores action. A wrong-direction predictor
is action-dependent but causally incorrect. An action-agnostic success prior predicts the
same attractive outcome for every candidate and therefore cannot rank them.

The toolkit makes three practical choices. First, an adapter declares whether it predicts
pixels, states, latents, actions, candidate scores, or world features. Unsupported metrics
are skipped with a reason. Second, uncertainty resamples complete contexts rather than
correlated branches or frames. Third, reports retain metric profiles and negative results;
WAMProbe does not publish a single total score.

\section{Related Evaluation Problems}

WAMs cover several interfaces, including imagine-then-execute systems, joint video-action
models, feature-conditioned policies, and action predictors trained with future-prediction
objectives. Their outputs are not interchangeable. Dynamic-consistency work argues for
checking whether predicted state changes agree with actions and warns that static
backgrounds can make failed trajectories appear consistent \cite{ruan2026compatible}.
Adaptive execution work identifies a separate deployment problem: a fixed action chunk
can continue after the observed state has diverged from the imagined future
\cite{wang2026trust}. WAMProbe addresses these concerns with action branches, explicit
no-op and transition-magnitude diagnostics, prefix execution, and replanning.

LIBERO provides task suites, simulator states, and standardized manipulation settings for
lifelong robot learning \cite{liu2023libero}. WAMProbe uses pinned LIBERO tasks as a paired
data-generation tier, not as a new policy leaderboard. The current real-model integration
targets the released StarWAM observation-to-action path \cite{pan2026starwam}. Because
that interface does not accept candidate actions or emit action-conditioned future states
or videos, candidate-action and future-fidelity metrics are structurally unavailable for
this adapter.

\section{Counterfactual Evaluation}

Let $c$ denote a context containing the observation, task, initial simulator state, and
goal. A candidate generator provides actions $\mathcal{A}(c)=\{a_1,\ldots,a_K\}$. For each
candidate, an adapter predicts a future $\hat{\tau}_i$, while a simulator or paired dataset
provides the reference future $\tau_i$. WAMProbe stores every branch under the same context
identifier and verifies that simulator branches begin from an identical snapshot.

\subsection{Causal and dynamics profile}

Action Dependence measures separation among predicted branch endpoints relative to the
separation among reference endpoints. It detects response to action but not correctness.
Counterfactual Direction compares predicted and true displacement directions. State ADE
and FDE measure trajectory and terminal error after semantic state alignment. No-op
fidelity checks the explicit no-op reference; it does not assume that a passive world,
such as a falling object, should freeze.

The use of separate measures is important. If $\hat{\tau}(a)$ follows the negative of the
true action effect, its branches remain well separated and Action Dependence can be high,
while direction and state error correctly identify the failure.

\subsection{Candidate ranking and regret}

A fixed task scorer $R(c,\tau)$ converts each future into a candidate value. Candidate
Ranking Correlation (CRC) reports Spearman correlation, Kendall $\tau_b$, NDCG, and
pairwise preference accuracy with explicit tie handling. Top-1 Regret is

\begin{equation}
\operatorname{Regret}(c)=
\max_{a_i\in\mathcal{A}(c)}R(c,\tau_i)
- R(c,\tau_{\hat\imath}),
\quad
\hat\imath=\arg\max_i R(c,\hat\tau_i).
\end{equation}

This definition keeps prediction and control semantics separate: the adapter predicts a
future, a fixed scorer values it, and regret is evaluated with the true branch outcome.

\subsection{Closed-loop utility}

The minimal closed-loop evaluator repeats four operations: score the legal candidate set,
select the highest-scoring future, execute only a prefix under true dynamics, and construct
the next context from the resulting state. The scoring prefix is capped by the remaining
episode budget. Without this cap, an exact fixed-horizon model can be penalized for a
terminal state beyond the steps that will actually be executed. The report records the
candidate order, execution prefix, total control steps, selected action sequence, final
return, success, and gap to a greedy simulator-future scorer for every context.

\section{Benchmarks and Implementation}

The dependency-free tier contains PointMass-2D, BlockPush-2D, and Gripper-Catch. PointMass
provides exact planar motion. BlockPush adds a pre-contact phase followed by directed
object motion. Gripper-Catch requires spatial alignment and an explicit close command
before a falling object becomes attached. The latter two benchmarks expose exact state
and a deterministic $64\times64$ RGB observation.

LIBERO-CF-Mini fixes four tasks spanning spatial, object, articulated-goal, and long-horizon
families. Each context branches into no-op, positive and negative end-effector motion, and
gripper closure for eight steps from the same simulator snapshot. The snapshot includes
MuJoCo integration state and Python-side controller, observable, gripper, clock, and random
state. Generation verifies two independent restores, a repeated branch, and forward versus
reverse branch order.

The package contains deterministic JSONL intervention data, content-addressed prediction
caches with payload checksums, context-level bootstrap intervals, exact-context paired
comparisons, and JSON, Markdown, and HTML reports. Python 3.11--3.13 continuous integration
runs formatting, linting, strict type checking, tests, and an 85\% coverage floor.

\section{Experiments}

\subsection{Reference baseline sanity checks}

Table~\ref{tab:pointmass} shows the PointMass profile. Oracle dynamics attain exact state
prediction and ranking. The wrong-direction baseline has maximal Action Dependence but
direction score $-1$, CRC $-0.686$, and the largest regret. Copy-last and action-agnostic
baselines have zero Action Dependence and zero CRC. BlockPush and Gripper-Catch reproduce
the same qualitative ordering while testing contact and attachment semantics.

\begin{table}[t]
\centering
\small
\caption{PointMass-2D reference profile over 12 shared contexts. Higher ADS, direction,
and CRC are better; lower FDE and regret are better.}
\label{tab:pointmass}
\begin{tabular}{lrrrrr}
\toprule
Model & ADS & Direction & FDE & CRC & Regret \\
\midrule
Oracle & 1.000 & 1.000 & 0.000 & 1.000 & 0.000 \\
Noisy linear & 0.989 & 0.997 & 0.093 & 0.977 & 0.000 \\
Copy last & 0.000 & 0.000 & 0.800 & 0.000 & 1.000 \\
Wrong direction & 1.000 & -1.000 & 1.600 & -0.686 & 2.000 \\
Action agnostic & 0.000 & 0.000 & 1.166 & 0.000 & 1.000 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Paired LIBERO data generation}

The LIBERO-CF-Mini pilot contains four tasks, four branches per task, eight post-action
steps, and two camera views. All independent restore, repeat, and branch-order checks pass;
the maximum integration-state error is 0.0. A second complete invocation validates the
snapshot sidecars, structured records, and every PNG by size and SHA256, then returns four
cache hits without opening the simulator. All diagnostic branch returns and sparse success
values are zero. This tier therefore validates paired data generation and branch
separation, not policy quality.

\subsection{StarWAM action prediction and execution}

The StarWAM matrix contains four tasks, three diffusion seeds, and action NFE values 1, 4,
and 8, for 36 predictions. Every inference succeeds, a repeated run obtains 36 verified
cache hits, and all 36 action chunks execute from their paired LIBERO snapshots. Mean
latencies are 0.780, 0.972, and 1.216 seconds for NFE 1, 4, and 8; peak allocated GPU memory
is approximately 11.39 GiB. Mean end-effector displacement for executed prefixes 8, 16,
and 32 is 0.1716, 0.1121, and 0.0000 meters. Longer execution is not monotonically more
useful in this diagnostic. Sparse return and success remain zero for every sample, so a
success correlation is undefined and no LIBERO task-solving claim is made.

\subsection{Video fidelity is not control value}

We evaluate six profiles on 12 BlockPush and 12 Gripper-Catch contexts. An
appearance-corrupted oracle inverts every predicted RGB value but preserves the exact
state future. It retains FDE 0, CRC 1, and regret 0 while obtaining PSNR 0.588 and 0.586 dB
on the two benchmarks. PSNR and regret ordering disagree on 3 of 9 and 5 of 9 comparable
profile pairs. The Pearson association between PSNR and regret is approximately $-0.16$
in each benchmark profile. This constructed counterexample does not imply that appearance
is irrelevant. It establishes the narrower claim that PSNR and a whole-frame global SSIM
diagnostic cannot substitute for state accuracy, ranking, or regret.

\subsection{Offline ranking and analytic closed-loop return}

Table~\ref{tab:closedloop} reports score-execute-observe evaluation with one executed step
per replan and 1,000 context-block bootstrap resamples. Oracle and noisy future scorers
solve BlockPush; the noisy scorer reaches 91.7\% exact-reference success on Gripper-Catch
with mean return 0.9967. Copy-last, wrong-direction, and action-agnostic future scorers have
zero success on both tasks. CRC and mean closed-loop return have Pearson association 0.9855
for BlockPush and 1.0000 for Gripper-Catch across the five future scorers. Regret and return
have association $-1.0000$ in both cases.

\begin{table}[t]
\centering
\small
\caption{Analytic closed-loop results over 12 contexts. Correlations use only the five
future-scorer rows; they are descriptive rather than population estimates.}
\label{tab:closedloop}
\begin{tabular}{llrrr}
\toprule
Task & Future scorer & Offline CRC & Mean return & Success \\
\midrule
BlockPush & Oracle & 1.000 & 0.000 & 1.000 \\
BlockPush & Noisy & 1.000 & 0.000 & 1.000 \\
BlockPush & Copy last & 0.000 & -0.300 & 0.000 \\
BlockPush & Wrong direction & -0.250 & -0.300 & 0.000 \\
BlockPush & Action agnostic & 0.000 & -0.300 & 0.000 \\
Gripper-Catch & Oracle & 1.000 & 1.000 & 1.000 \\
Gripper-Catch & Noisy & 1.000 & 0.997 & 0.917 \\
Gripper-Catch & Copy last & 0.000 & 0.000 & 0.000 \\
Gripper-Catch & Wrong direction & 0.000 & 0.000 & 0.000 \\
Gripper-Catch & Action agnostic & 0.000 & 0.000 & 0.000 \\
\bottomrule
\end{tabular}
\end{table}

The result confirms the evaluator path from offline ranking to repeated execution on two
controlled tasks. It does not show that a learned WAM improves control: the five profiles
are intentionally separated analytic baselines, three weak profiles tie on return, and a
fixed action-only policy also reaches the task optimum.

\section{Reproducibility and Release Discipline}

Model weights, raw RGB, simulator states, and generated predictions remain outside Git.
Committed manifests pin upstream revisions and configuration hashes, while reports retain
the hashes of local evidence indexes. A release-candidate command fixes the source date to
the commit timestamp, builds wheel and source distributions twice, requires byte-identical
SHA256 values, rejects unsafe or forbidden archive paths, checks committed evidence, and
installs the wheel without network access in a fresh virtual environment. Public release
is separate from candidate construction and requires independent reproduction and
maintainer approval.

\section{Limitations}

The analytic benchmarks have known low-dimensional dynamics and cannot establish visual,
physical, or task-distribution transfer. The closed-loop correlations contain five
designed profiles and should not be interpreted as statistical evidence over trained
models. LIBERO-CF-Mini has one initial state per task family and diagnostic actions rather
than expert candidates. The StarWAM slice exposes actions but no action-conditioned future
artifact, preventing the central candidate-future metrics on that model. The real-model
matrix contains short open-loop executions with constant sparse return. A paper-level
claim about WAM control utility requires more initial states and at least one adapter that
can score the same candidate set through predicted futures.

The current whole-frame global SSIM implementation is a dependency-free diagnostic, not
the standard windowed metric. Pixel metrics are secondary. The toolkit also does not
provide real-robot safety certification, and it does not normalize all robot action spaces
into a universal representation.

\section{Conclusion}

WAMProbe turns action conditioning from an informal expectation into a paired evaluation
contract. Its current evidence supports three limited conclusions: reference metrics
separate known causal failures, pixel fidelity cannot replace control-grounded evaluation,
and offline candidate ranking tracks closed-loop return in two analytic sanity tasks. The
real StarWAM experiment establishes a reproducible action-prediction and execution path but
does not yet establish task success or candidate-future utility. Preserving these boundaries
is central to the project: unsupported metrics are skipped, negative results remain visible,
and every public claim is linked to structured evidence.

\bibliographystyle{plain}
\bibliography{references}

\end{document}
